---
title: "R Data Science Course 9: Choose Your Own Project"
author: "James Lo"
date: '2022-06-09'
output: 
  pdf_document:
    latex_engine: xelatex
    df_print: kable
    toc : TRUE
    toc_depth: 3
urlcolor: blue
mainfont: Calibri Light
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Introduction

(words go here)

## Data Loading

```{r packages, warning = F}
if(!require(tidyverse)) install.packages("tidyverse",
                                         repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret",
                                     repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", 
                                          repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", 
                                     repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", 
                                      repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", 
                                         repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readr)) install.packages("readr", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", 
                                       repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(data.table)
library(dslabs)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(readxl)
# will add more packages as needed

```

```{r data_download, warning = FALSE}
uci_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/00602/DryBeanDataset.zip'

download.file(uci_url, 'DryBeanDataset.zip')
```
```{r data_unzip, warning = FALSE}
unzip('DryBeanDataset.zip')
```

```{r data_read, warning = FALSE}
# specify path
path <- paste0(getwd(), '/DryBeanDataset/Dry_Bean_Dataset.xlsx')
# read the first sheet (where the data is)
beans <- read_xlsx(path, sheet = 1)
```

## Exploratory Analysis

### Overview

```{r data_overview}
# overview of the data's size and format
str(beans)
```
The dataset contains 13611 entries, each representing a scanned bean. There are 12 metrics describing the dimensions of each bean, plus four 'shape factors' which are derived from a previous machine learning article on rice grains (cite).

### Checking for NA Values

```{r na_check}
which(is.na(beans))
```
As described on the UCI website, the dataset indeed contains no NA entries.

### Distribution of Outcome Classes

```{r num_classes}
# checking the distribution of classes (bean types) in the dataset
beans %>% group_by(Class) %>%
  summarize(count = n())
```
The dataset has a rather uneven distribution of the 7 bean varieties. The most frequently found type is Dermason at 3546 entries, while the least common is Bombay at 522 entries. This uneven distribution must be accounted for during model fitting and accuracy assessments.

### Distribution of Potential Predictor Variables Among Classes

Before models could be built, the distribution of values among differnt bean varieties were visualized with boxplots for all 16 possible predictor variables. This was done to understand the levels of variability between bean varieties that were present in each potential predictor, and select the most informative predictors for models that perform best with lower-dimensional data.

#### Area

```{r Area_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Area > (quantile(Area, 0.75)) + 1.5 * IQR(Area) | 
           Area < (quantile(Area, 0.25)) - 1.5 * IQR(Area))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Area), y = Area)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Areas')
```
The area IQR for Bombay was distinctively higher than those of all other varieties. Heavily overlapping IQRs exist in Cali and Barbunya, as well as Sira and Seker. Dermason, Seker, and Sira all have much narrower IQRs compared to the other varieties, despite having very close distribution ranges. Most outliers were found on the upper ends of their respective distributions.

#### Perimeter

```{r Perimeter_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Perimeter > (quantile(Perimeter, 0.75)) + 1.5 * IQR(Perimeter) | 
           Perimeter < (quantile(Perimeter, 0.25)) - 1.5 * IQR(Perimeter))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Perimeter), y = Perimeter)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Perimeters')
```
Similarly to the Area metric, Bombay had a distinctively higher range of Perimeter lengths than the other varieties. The IQR for Cali existed entirely within that of Barbunya, while the IQRs for Dermason and Seker overlapped. The IQRs of perimeter lengths for smaller bean varieties were less tight than seen in the Area graph, but also somewhat more distinct relative to the scale of the y-axis.

#### Major Axis Length

```{r MajorAxisLength_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(MajorAxisLength > (quantile(MajorAxisLength, 0.75)) + 1.5 * IQR(MajorAxisLength) | 
           MajorAxisLength < (quantile(MajorAxisLength, 0.25)) - 1.5 * IQR(MajorAxisLength))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, MajorAxisLength), y = MajorAxisLength)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Major Axis Lengths')
```
For Major Axis Length, Bombay once again had an IQR that was distinctively larger than all others. Barbunya/Horoz and Dermason/Seker had heavily overlapping IQRs, while Cali had a smaller IQR overlap with Horoz/Barbunya. Seker and Sira had more outliers on the high end of their ranges, while Horoz had more on the low end.

#### Minor Axis Length

```{r MinorAxisLength_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(MinorAxisLength > (quantile(MinorAxisLength, 0.75)) + 1.5 * IQR(MinorAxisLength) | 
           MinorAxisLength < (quantile(MinorAxisLength, 0.25)) - 1.5 * IQR(MinorAxisLength))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, MinorAxisLength), y = MinorAxisLength)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Minor Axis Lengths')
```

The distribution of Minor Axis Lengths was similar to that of Major Axis lengths, but with a few differences: here the heaviest overlap was between Cali and Barbunya, followed by Horoz/Sira and Sira/Seker. Also, Dermason had a partial IQR overlap with Horoz rather than a near-complete one with Seker. Most outliers were on the high ends of their respective groups. Cali, Seker, and Horoz had more outliers on th high ends of their respective groups, while Sira had more on the low end.

#### Aspect Ratio

```{r AspectRation_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(AspectRation > (quantile(AspectRation, 0.75)) + 1.5 * IQR(AspectRation) | 
           AspectRation < (quantile(AspectRation, 0.25)) - 1.5 * IQR(AspectRation))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, AspectRation), y =AspectRation)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Aspect Ratios')
```

The Aspect Ratio distributions differed from the previous potential predictors in that Bombay no longer had the most distinct distribution of values. Instead, Horoz and Cali had IQRs distinctively higher than others, while Seker had a lower IQR. The remaining varieties (Dermason, Barbunya, Sira, Bombay) all had heavily overlapping IQRs. Horoz had more outliers on the bottom end of its distribution, while Sira, Dermason, and Seker had more on their high ends.

#### Eccentricity

```{r Eccentricity_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Eccentricity > (quantile(Eccentricity, 0.75)) + 1.5 * IQR(Eccentricity) | 
           Eccentricity < (quantile(Eccentricity, 0.25)) - 1.5 * IQR(Eccentricity))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Eccentricity), y = Eccentricity)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Eccentricities')
```

The Eccentricity distributions were similar to that of Aspect Ratios, but with the two varieties having higher IQRs (Horoz, Cali) deviating from the others less while the one with lower IQR (Seker) deviating more. Most outliers were on the low ends of their respective ranges.

#### Convex Area

```{r ConvexArea_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(ConvexArea > (quantile(ConvexArea, 0.75)) + 1.5 * IQR(ConvexArea) | 
           ConvexArea < (quantile(ConvexArea, 0.25)) - 1.5 * IQR(ConvexArea))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, ConvexArea), y = ConvexArea)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Convex Areas')

```

With the Covex Area Sizes, a distribution somewhat similar to the Area, Perimeter, and Axis Length distributions could be seen again. Bombay had a distinctively high IQR of Convex Areas, while Dermason had the lowest range. Cali and Barbunya had heavily overlapping IQRs, with less overlap existing among the other varieties. The narrowest IQRs belonged to Seker and Sira. Outliers were biased towards the top end for Bombay, Cali, Barbunya, Sira, and Seker.

#### Equivalent Diameter

```{r EquivDiameter_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(EquivDiameter > (quantile(EquivDiameter, 0.75)) + 1.5 * IQR(EquivDiameter) | 
           EquivDiameter < (quantile(EquivDiameter, 0.25)) - 1.5 * IQR(EquivDiameter))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, EquivDiameter), y = EquivDiameter)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Equivalent Diameters')
```

The distributions of Equivalent Diameters were fairly similar to that of Convex Areas. Once again, Cali and Barbunya had heavy overlaps, with Bombay having a distinctly higher IQR. Cali and Seker had more outliers on the top ends of their distributions, while Horoz had more on the bottom end.

#### Extent

```{r Extent_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Extent > (quantile(Extent, 0.75)) + 1.5 * IQR(Extent) | 
           Extent < (quantile(Extent, 0.25)) - 1.5 * IQR(Extent))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Extent), y = Extent)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Extents')
```

The distribution of Extent Ratios exhibited heavy IQR overlaps between all bean varieties. Among the three groups with outliers (Cali, Seker, Bombay), all had more on the bottom ends of their respective distributions.

#### Solidity

```{r Solidity_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Solidity > (quantile(Solidity, 0.75)) + 1.5 * IQR(Solidity) | 
           Solidity < (quantile(Solidity, 0.25)) - 1.5 * IQR(Solidity))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Solidity), y = Solidity)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Solidities')
```

The distribution of Solidity values exhibited heavy IQR overlaps between nearly all bean varieties. Also, every bean variety had many outliers (values outside the boxplot whiskers) on the lower end of their respective distributions. Outliers were abundant, and located on the low ends of every group's distribution.

#### Roundness

```{r roundness_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(roundness > (quantile(roundness, 0.75)) + 1.5 * IQR(roundness) | 
           roundness < (quantile(roundness, 0.25)) - 1.5 * IQR(roundness))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, roundness), y = roundness)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Roundnesses')
```

The ranking of median Roundness values were similar to that of that of Solidity, but with the lowest three (Cali, Barnunya, Horoz) switched. Less overlap existed than in Solidity (except for heavy overlap Horoz and Barbunya), but many outliers outside the whiskers existed for all varieties. Outliers were abundant, and biased towards the low end of the range for all groups except Bombay.

#### Compactness

```{r Compactness_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(Compactness > (quantile(Compactness, 0.75)) + 1.5 * IQR(Compactness) | 
           Compactness < (quantile(Compactness, 0.25)) - 1.5 * IQR(Compactness))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, Compactness), y = Compactness)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Compactness Values')
```

The distribution of Compactness values had heavy overlaps between Bombay, Sira, Barbunya, and Dermason. Seker had the highest range of Compactness, while Horoz had the lowest. All varieties had many outliers, mostly on the high ends of the distributions. Seker had most of its outliers on the low end of its range; the opposite was true for Bombay, Cali, and Horoz.

#### Shape Factor 1

```{r ShapeFactor1_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(ShapeFactor1 > (quantile(ShapeFactor1, 0.75)) + 1.5 * IQR(ShapeFactor1) | 
           ShapeFactor1 < (quantile(ShapeFactor1, 0.25)) - 1.5 * IQR(ShapeFactor1))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, ShapeFactor1), y = ShapeFactor1)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 1')
```

The distributions of Shape Factor 1 values featured heavy overlap between Barbunya/Cali, and relatively less overlap between Seker/Sira/Horoz. Dermason had the highest range of Shape Factor 1, and Bombay had the lowest. All groups had more outliers on the top ends of their ranges except for Seker (more on the low end) and Bombay (two outliers on the low end only).

#### Shape Factor 2

```{r ShapeFactor2_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(ShapeFactor2 > (quantile(ShapeFactor2, 0.75)) + 1.5 * IQR(ShapeFactor2) | 
           ShapeFactor2 < (quantile(ShapeFactor2, 0.25)) - 1.5 * IQR(ShapeFactor2))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, ShapeFactor2), y = ShapeFactor2)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 2')
```

The distributions of Shape Factor 2 values had relatively less IQR overlap compared to most other variables examined, with the 4 highest Shape Factor 2 ranges (Seker, Dermason, Sira, Barbunya) all having non-overlapping IRRs. Horoz and Cali had a large IQR overlap, but neither had overlapping IQRs with Bombay. However, outliers on the high ends of the ranges were common. With the exception of Seker, all groups had more outliers on the top ends of their ranges.

#### Shape Factor 3

```{r ShapeFactor3_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(ShapeFactor3 > (quantile(ShapeFactor3, 0.75)) + 1.5 * IQR(ShapeFactor3) | 
           ShapeFactor3 < (quantile(ShapeFactor3, 0.25)) - 1.5 * IQR(ShapeFactor3))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, ShapeFactor3), y = ShapeFactor3)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 3')
```

The distributions of Shape Factor 3 values had heavy overlaps in Bombay, Sira, Barbunya, and Dermason. Seker had a distinctively higher IQR than the other varieties but also many low outliers; the opposite was true for Horoz. Except for Seker, all groups had more outliers on the low ends of their ranges.

#### Shape Factor 4

```{r ShapeFactor4_boxplot}
# separately mark outliers
outliers <-
  beans %>%
  # do the following for each Class
  group_by(Class) %>%
  # filter for outliers; whiskers are 1.5 times IQR, while boxes go from Q1 to Q3
  filter(ShapeFactor4 > (quantile(ShapeFactor4, 0.75)) + 1.5 * IQR(ShapeFactor4) | 
           ShapeFactor4 < (quantile(ShapeFactor4, 0.25)) - 1.5 * IQR(ShapeFactor4))

beans %>%
  # plot with jittering outliers
  ggplot(aes(x = reorder(Class, ShapeFactor4), y = ShapeFactor4)) +
  # outliers plotted separately in geom_point()
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data = outliers) + 
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 4')
```

The distributions for Shape Factor 4 exhibited heavy overlaps between nearly all varieties, with the possible exception of Seker. In addition, every variety had many outliers on the low ends of their respective distributions. All groups had outliers on the low ends of their distributions; Horoz had an exceptionally long range of outlier values.

### Combining Predictors for Scatterplots

Some common classification models, such as KNN, do not perform efficiently with high-dimentional data. As such, it would be constructive to select a pair of "compatible" predictor variables for those models. The goal is to have each variable contain inter-group (between bean varieties) variances that the other does not have sufficient amounts of. Also, the variables should be independent, lacking in redundancy of measurements. For instance, if one variable in the pair is already derived from bean lengths, the other variable should not also contain bean lengths in its formula. In addition, each of the two variables should have as much distance (or failing that, the least amount of overlap) between its group IQRs as possible, with the least amount (and skew) of outliers. Note that while some variables (like Area and Convex Area) exist on a far larger scale (at least in numbers alone) than others, normalization would make relative distances along the y-axis (as seen on the boxplots) more relevant than raw numerical differences.

To identify promising combinations, the variables were roughly grouped into six categories, reflecting broad features in their boxplots:

* Group 0 ("don't bother"): Variables whose boxplots displayed heavily overlapping IQRs and outliers that are visibly more numerous (compared to the other variables in this dataset) and/or heavily skewed in distributions. Extent, Solidity, and Shape Factor 4 were placed under this category.

* Group 1 ("Bombay high, Dermason low"): Variables where Bombay had the highest IQR (usually by a relatively large margin), and Dermason the lowest. Other bean varieties tended to have less separate or even overlapping IQRs. Area, Perimeter, Major Axis, Minor Axis, Convex Area, and Equivalent Diameter all fell under this category.

* Group 2 ("Horoz high, Seker low"): Variables where Horoz had the highest IQR, and Seker the lowest. Again, other groups tended to have less separate or overlapping IQRs. Aspect Ratio and Eccentricity fell under this cetegory.

* Group 3 ("Seker high, Horoz low"): The opposite of group 2.  Roundness and Compactness fell under this category. Note that as described in the original paper, Roundness and Compactness are both measures of how round a bean is, albeit calculated by different formulae. Both are also, by definition, the opposite metric of Aspect Ratio (which measures how elongated a bean is through the ratio between length/Major Axis Length and width/Minor Axis Width).

* Group 4 ("Dermason high, Bombay low"): The opposite of group 1, at least in terms of the defining feature (arbitrarily assigned for quick grouping). Shape Factor 1 was the sole variable assigned to this category.

* Group 5 ("Seker high, Bombay low"): Shape Factor 2 was the only variable assigned under this group.

Among group 1 variables, Perimeter and Equivalent Diameter appeared to be the most suitable choices, as they both had relatively less overlaps between the "middle groups" representing varieties other than Bombay and Dermason. Perimeter had large overlaps in Cali/Barbunya and Seker/Dermason, while Equivalent Diameter had large overlaps in Cali/Barbunya and Sira/Seker. Observing the boxplots, Aspect Ratio, Compactness, and Shape Factor 3 were selected as the best complements due to having higher variances lacking in the group 1 choices, in addition to relatively less numerous and skewed outliers. Equivalent Diameter is incompatible with Compactness (which is derived form Equivalent Diameter) and Shape Factor 3 (which is derived from Area, like ED), while Perimeter is compatible with Aspect Ratio, Compactness, and Shape Factor 3. Included below are four scatterplots to further visualize the variables' ability to model bean varieties as distinct clusters of data points.

```{r Perim_AR}
```

```{r Perim_Comp}
```

```{r Perim_SF3}
```

```{r ED_AR}
```

# Methods

# Results

# Conclusion

# Citation
