---
title: "R Data Science Course 9: Choose Your Own Project"
author: "James Lo"
date: '2022-06-09'
output: 
  pdf_document:
    latex_engine: xelatex
    df_print: kable
    toc : TRUE
    toc_depth: 3
urlcolor: blue
mainfont: Calibri Light
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Introduction

(words go here)

## Data Loading

```{r packages, warning = F}
if(!require(tidyverse)) install.packages("tidyverse",
                                         repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret",
                                     repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", 
                                          repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", 
                                     repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", 
                                      repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", 
                                         repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readr)) install.packages("readr", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", 
                                       repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(data.table)
library(dslabs)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(readxl)
# will add more packages as needed

```

```{r data_download, warning = FALSE}
uci_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/00602/DryBeanDataset.zip'

download.file(uci_url, 'DryBeanDataset.zip')
```
```{r data_unzip, warning = FALSE}
unzip('DryBeanDataset.zip')
```

```{r data_read, warning = FALSE}
# specify path
path <- paste0(getwd(), '/DryBeanDataset/Dry_Bean_Dataset.xlsx')
# read the first sheet (where the data is)
beans <- read_xlsx(path, sheet = 1)
```

## Exploratory Analysis

### Overview

```{r data_overview}
# overview of the data's size and format
str(beans)
```
The dataset contains 13611 entries, each representing a scanned bean. There are 12 metrics describing the dimensions of each bean, plus four 'shape factors' which are derived from a previous machine learning article on rice grains (cite).

### Checking for NA Values

```{r na_check}
which(is.na(beans))
```
As described on the UCI website, the dataset indeed contains no NA entries.

### Distribution of Outcome Classes

```{r num_classes}
# checking the distribution of classes (bean types) in the dataset
beans %>% group_by(Class) %>%
  summarize(count = n())
```
The dataset has a rather uneven distribution of the 7 bean varieties. The most frequently found type is Dermason at 3546 entries, while the least common is Bombay at 522 entries. This uneven distribution must be accounted for during model fitting and accuracy assessments.

### Distribution of Potential Predictor Variables Among Classes

(find a combination of 2 variables that combine to separate the varieties into distinct groups, do color-coded scatterplots to confirm good choice)
(should finish this part by the 19th; 17th for describing variables, 18th for plotting, 19th for troubleshooting)
(20th, 21th will be for outline of methodologies)
(22-24th will be for trying first model: log reg)
(25th will be an off day for reasons)
(26-27th will be more first model or beginnings of second model: the KNN)
(don't forget to have a life in the meantime, and not spend your whole day like today the 16th)

#### Area

```{r Area_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Area), y = Area)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Areas')
```
The area IQR for Bombay was distinctively higher than those of all other varieties. Heavily overlapping IQRs exist in Cali and Barbunya, as well as Sira and Seker. Dermason, Seker, and Sira all have much narrower IQRs compared to the other varieties, despite having very close distribution ranges.


```{r Perimeter_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Perimeter), y = Perimeter)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Perimeter Sizes')
```
Similarly to the Area metric, Bombay had a distinctively higher range of Perimeter lengths than the other varieties. The IQR for Cali existed entirely within that of Barbunya, while the IQRs for Dermason and Seker overlapped. The IQRs of perimeter lengths for smaller bean varieties were less tight than seen in the Area graph, but also somewhat more distinct relative to the scale of the y-axis.

```{r MajorAxisLength_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, MajorAxisLength), y = MajorAxisLength)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Major Axis Lengths')
```
For Major Axis Length, Bombay once again had an IQR that was distinctively larger than all others. Barbunya/Horoz and Dermason/Seker had heavily overlapping IQRs, while Cali had a smaller IQR overlap with Horoz/Barbunya.
```{r MinorAxisLength_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, MinorAxisLength), y = MinorAxisLength)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Minor Axis Lengths')
```
The distribution of Minor Axis Lengths was similar to that of Major Axis lengths, but with a few differences: here the heaviest overlap was between Cali and Barbunya, followed by Horoz/Sira and Sira/Seker. Also, Dermason had a partial IQR overlap with Horoz rather than a near-complete one with Seker.

```{r Aspectratio_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, AspectRation), y = AspectRation)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Aspect Ratios')
```

The Aspect Ratio distributions differed from the previous potential predictors in that Bombay no longer had the most distinct distribution of values. Instead, Horoz and Cali had IQRs distinctively higher than others, while Seker had a lower IQR. The remaining varieties (Dermason, Barbunya, Sira, Bombay) all had heavily overlapping IQRs.

```{r Eccentricity_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Eccentricity), y = Eccentricity)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Eccentricity Values')
```

The Eccentricity distributions were similar to that of Aspect Ratios, but with the two varieties having higher IQRs (Horoz, Cali) deviating from the others less while the one with lower IQR (Seker) deviating more.

```{r ConvexArea_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ConvexArea), y = ConvexArea)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Convex Area Sizes')
```



```{r EquivDiameter_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, EquivDiameter), y = EquivDiameter)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Equivalent Diameters')
```
```{r Extent_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Extent), y = Extent)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Extent Ratios')
```
```{r Solidity_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Solidity), y = Solidity)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Solidity Ratios')
```
```{r roundness_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, roundness), y = roundness)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Roundness Values')
```
```{r Compactness_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Compactness), y = Compactness)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Compactness Values')
```
```{r ShapeFactor1_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor1), y = ShapeFactor1)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 1 Values')
```
```{r ShapeFactor2_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor2), y = ShapeFactor2)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 2 Values')
```
```{r ShapeFactor3_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor3), y = ShapeFactor3)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 3 Values')
```
```{r ShapeFactor4_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor4), y = ShapeFactor4)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 4 Values')
```

# Methods

# Results

# Conclusion

# Citation
