---
title: "R Data Science Course 9: Choose Your Own Project"
author: "James Lo"
date: '2022-06-09'
output: 
  pdf_document:
    latex_engine: xelatex
    df_print: kable
    toc : TRUE
    toc_depth: 3
urlcolor: blue
mainfont: Calibri Light
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Introduction

```{r packages, warning = F}
if(!require(tidyverse)) install.packages("tidyverse",
                                         repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret",
                                     repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", 
                                          repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", 
                                     repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", 
                                      repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", 
                                         repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readr)) install.packages("readr", 
                                       repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", 
                                       repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(data.table)
library(dslabs)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(readxl)
# will add more packages as needed

```

```{r data_download, warning = FALSE}
uci_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/00602/DryBeanDataset.zip'

download.file(uci_url, 'DryBeanDataset.zip')
```
```{r data_unzip, warning = FALSE}
unzip('DryBeanDataset.zip')
```

```{r data_read, warning = FALSE}
# specify path
path <- paste0(getwd(), '/DryBeanDataset/Dry_Bean_Dataset.xlsx')
# read the first sheet (where the data is)
beans <- read_xlsx(path, sheet = 1)
```

```{r data_overview}

# overview of the data's size and format
str(beans)
```
The dataset contains 13611 entries, each representing a scanned bean. There are 12 metrics describing the dimensions of each bean, plus four 'shape factors' which are derived from a previous machine learning article on rice grains (cite).

```{r na_check}
which(is.na(beans))
```
As described on the UCI website, the dataset indeed contains no NA entries.

```{r num_classes}
# checking the distribution of classes (bean types) in the dataset
beans %>% group_by(Class) %>%
  summarize(count = n())
```
The dataset has a rather uneven distribution of the 7 bean varieties. The most frequently found type is Dermason at 3546 entries, while the least common is Bombay at 522 entries. This uneven distribution must be accounted for during model fitting and accuracy assessments.

```{r Area_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Area), y = Area)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Areas')
```
The area IQR for Bombay was distinctively higher than those of all other varieties. Heavily overlapping IQRs exist in Cali and Barbunya, as well as Sira and Seker. Dermason, Seker, and Sira all have much narrower IQRs compared to the other varieties, despite having very close distribution ranges.


```{r Perimeter_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Perimeter), y = Perimeter)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Perimeter Sizes')
```

```{r MajorAxisLength_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, MajorAxisLength), y = MajorAxisLength)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Major Axis Lengths')
```
```{r MinorAxisLength_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, MinorAxisLength), y = MinorAxisLength)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Minor Axis Lengths')
```
```{r Aspectratio_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, AspectRation), y = AspectRation)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Aspect Ratios')
```
```{r Eccentricity_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Eccentricity), y = Eccentricity)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Eccentricity Values')
```
```{r ConvexArea_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ConvexArea), y = ConvexArea)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Convex Area Sizes')
```
```{r EquivDiameter_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, EquivDiameter), y = EquivDiameter)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Equivalent Diameters')
```
```{r Extent_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Extent), y = Extent)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Extent Ratios')
```
```{r Solidity_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Solidity), y = Solidity)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Solidity Ratios')
```
```{r roundness_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, roundness), y = roundness)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Roundness Values')
```
```{r Compactness_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, Compactness), y = Compactness)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Compactness Values')
```
```{r ShapeFactor1_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor1), y = ShapeFactor1)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 1 Values')
```
```{r ShapeFactor2_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor2), y = ShapeFactor2)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 2 Values')
```
```{r ShapeFactor3_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor3), y = ShapeFactor3)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 3 Values')
```
```{r ShapeFactor4_boxplot}
beans %>%
  ggplot(aes(x = reorder(Class, ShapeFactor4), y = ShapeFactor4)) +
  geom_boxplot() +
  xlab('Class') +
  ggtitle('Distribution of Shape Factor 4 Values')
```

# Methods

# Results

# Conclusion

# Citation
